Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-05-31 docs/submission/template.tex
collaborator comments

\begin{abstract}
Water bodies exchange mass and energy with the atmosphere and need to be accurately simulated in numerical weather prediction and climate modelling as they strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. All the non-ocean water (resolved and sub-grid lakes and coastal waters) are represented in the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF) model, by the Fresh-water Lake (FLake) parametrisation, which treats $\sim$ 1/3 of the land. It is a continuous enterprise to update the surface parametrization schemes and their input fields to better represent small-scale processes. It is, however, difficult to quickly determine both the accuracy of an updated parametrisation, and the added value gained for the purposes of numerical modelling. In this work a neural network regression model is developed to learn the mapping between surface data from ERA5 and MODIS satellite observations of skin temperature. Once trained, the tool can be used to verify the quality of upgrades of the land-surface parametrisation scheme since large differences between the model and the machine learning tool indicate incorrect values, which are most often caused by errors in the surface fields that are used as input to the parametrisation scheme. We deploy this model to determine the accuracy of recent upgrades to the FLake parametrisation, namely the improved permanent lake cover and the capacity to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature improves by 0.37 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 0.84 K. We also show that updates to the glacier cover improve further the prediction accuracy by 0.22 K.

@@

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
\noindent Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic, over saturated with CO$_2$, as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO2-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Huang2019,Peng2020,Franz2018}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

 \begin{figure*}
 	\subfloat[\label{fig:wasserstein_temperature}]{\includegraphics[width=0.48\textwidth]{model_prediction_error.png}} \hspace{1mm}
 	\subfloat[\label{fig:wasserstein_precipitation}]{\includegraphics[width=0.48\textwidth]{ERA_prediction_error.png}}
 	\caption{Mean absolute error (MAE, $\delta K$) of LST predictions for 2019 at ~31km resolution based on differences between (a) VESPER\_V15 (i.e. VESPER trained with V15 surface physiographic fields) and Aqua-MODIS LST, and (b) between ERA5 skin temperature and Aqua-MODIS LST. It can be seen that VESPER\_V15 managed to learn corrections over regions with complex surface fields such as the Himalayas (lots of orography) sub-Saharan Africa (lots of vegetation) and the Amazon Basin (lots of water + vegetation). } 
 	\label{fig:example_model}
 \end{figure*}

>>>>

\begin{abstract}
A purposely built deep learning algorithm for the Verification of Earth-System ParametERisation (VESPER) is used to assess recent upgrades of the global physiographic datasets underpinning the quality of the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF), which is used both in numerical weather prediction and climate reanalyses. A neural network regression model is trained to learn the mapping between the surface physiographic dataset plus the meteorology from ERA5, and the MODIS satellite skin temperature observations. Once trained, this tool is applied to rapidly assess the quality of upgrades of the land-surface scheme. Upgrades which improve the prediction accuracy of the machine learning tool indicate a reduction of the errors in the surface fields used as input to the surface parametrisation schemes. Conversely, incorrect specifications of the surface fields decrease the accuracy with which VESPER can make predictions.  We apply VESPER to assess the accuracy of recent upgrades of the permanent lake and glaciers covers as well as planned upgrades to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature improves by 0.37 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 0.84 K. We also show that updates to the glacier cover improve further the prediction accuracy by 0.22 K. We highlight how neural networks such as VESPER can assist the research and development of surface parametrizations and their input physiography to better represent Earth’s surface coupled processes in weather and climate models.

%Water bodies exchange mass and energy with the atmosphere and need to be accurately simulated in numerical weather prediction and climate modelling as they strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. All the non-ocean water (resolved and sub-grid lakes and coastal waters) are represented in the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF) model, by the Fresh-water Lake (FLake) parametrisation, which treats $\sim$ 1/3 of the land. It is a continuous enterprise to update the surface parametrization schemes and their input fields to better represent small-scale processes. It is, however, difficult to quickly determine both the accuracy of an updated parametrisation, and the added value gained for the purposes of numerical modelling. In this work a neural network regression model is developed to learn the mapping between surface data from ERA5 and MODIS satellite observations of skin temperature. Once trained, the tool can be used to verify the quality of upgrades of the land-surface parametrisation scheme since large differences between the model and the machine learning tool indicate incorrect values, which are most often caused by errors in the surface fields that are used as input to the parametrisation scheme. We deploy this model to determine the accuracy of recent upgrades to the FLake parametrisation, namely the improved permanent lake cover and the capacity to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature improves by 0.37 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 0.84 K. We also show that updates to the glacier cover improve further the prediction accuracy by 0.22 K.

@@

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
Accurate knowledge of the global surface physiography, including land, water and ice covers, and their characteristics, strongly determines the quality of surface and near-surface temperature simulations in weather and climate modelling. For instance, water bodies exchange mass and energy with the atmosphere and their thermal inertia strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic, over saturated with CO$_2$, as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO2-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Huang2019,Peng2020,Franz2018}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

 \begin{figure*}
 	\subfloat[\label{fig:wasserstein_temperature}]{\includegraphics[width=0.48\textwidth]{ERA_prediction_error.png}} \hspace{1mm}
 	\subfloat[\label{fig:wasserstein_precipitation}]{\includegraphics[width=0.48\textwidth]{model_prediction_error.png}}
 	\caption{Mean absolute error (MAE, $\delta K$) of LST predictions for 2019 at ~31km resolution based on differences between (a) ERA5 skin temperature and Aqua-MODIS LST and  (b) between VESPER\_V15 (i.e. VESPER trained with V15 surface physiographic fields) and Aqua-MODIS LST. It can be seen that VESPER\_V15 managed to learn corrections over regions with complex surface fields such as the Himalayas (lots of orography) sub-Saharan Africa (lots of vegetation) and the Amazon Basin (lots of water + vegetation). } 
 	\label{fig:example_model}
 \end{figure*}
